---
title: "PA1_template"
author: "Manfredi Ruggeri"
date: "2/8/2020"
output: html_document
---

## Reproducible Research: PA1

**Loading and preprocessing the data**  
```{r warning=FALSE}
#load libraries 
library(dplyr)
library(lubridate)
library(ggplot2)
#read data
act <- read.csv("activity.csv", header=TRUE)
#process data
act$date <- as.Date(as.character(act$date), "%Y-%m-%d") 
```
# What is mean total number of steps taken per day?
```{r warning=FALSE}
tb <- act %>% group_by(date) %>% summarise(steps=sum(steps, na.rm=TRUE))
```
  
**Make a histogram of the total number of steps**  
```{r}
hist(tb$steps, ylim=c(0,30), xlab="Steps", main ="Histogram of steps per day")
```
  
**Calculate and report the mean and median of the total number of steps taken per days **  

mean:
```{r}
print(mean(tb$steps))
```
median:
```{r}
print(median(tb$steps))
```
# What is the average daily activity pattern?


**Make a time series plot of steps averaged across all days**  


group data by interval and calculate the mean across groups
```{r warning=FALSE}
tb2 <- act %>% group_by(interval) %>% summarise(mean=mean(steps, na.rm=TRUE))
plot(tb2$interval, tb2$mean, type="l", xlab="Interval", ylab="Steps averaged across days")
grid()
```
  

**Which interval contains the maximum number of steps?**  
```{r}
m <- max(tb2$mean)
ans <- tb2 %>% filter(mean==m)
print(ans[1,1])
```
# Imputing missing values  
**Calculate and report the total number of missing values in the dataset**  
```{r}
naact <- act %>% filter(is.na(steps) | is.na(date) | is.na(interval))
print(dim(naact))
```
So we have 2304 rows affected by NAs  
**Devise a strategy for filling in all of the missing values in the dataset**    
First of all we need to find where NAs are  
```{r warning=FALSE}
naSteps <- act %>% filter(is.na(steps))
naDate <- act %>% filter(is.na(date))
nainterval <- act %>% filter(is.na(interval))
print(dim(naSteps))
print(dim(naDate))
print(dim(nainterval))
```
Luckly only steps variable has NAs.    
Now we fill missing data in variable steps with the steps averaged across 5-minute interval    
**Create a new dataset with the missing data filled in**   
```{r}
##completed rows in dataframe act
notnaact <- act %>%filter(!is.na(steps))
#fill missing data
for(i in 1:nrow(naSteps)) {
      tmp1 <- filter(tb2, interval== naSteps[i,3])
      naSteps[i, 1]<- tmp1[1,2]
}
fillAct <- rbind(naSteps, notnaact)
```
  
**Make an histogram of the total numbers of steps taken per day**    
```{r}
tb3 <- fillAct %>% group_by(date) %>% summarise(steps=sum(steps))
hist(tb3$steps, ylim=c(0,40), xlab="Steps", main ="New Histogram of steps per day")
```
  
**Calculate and report the mean and median total number of steps taken for day**  
mean:    
```{r}
print(mean(tb3$steps))
```
median:  
```{r}
print(median(tb3$steps))
```
After imputing missing data, mean and median are identical    
# Are there differences in activity patterns between weekdays and weekends?  
**Create a new factor variable in dataset with two levels "weekday" and "weekends"**    
```{r warning=FALSE}
fillAct$dType <- ifelse(wday(fillAct$date)>5, "weekend", "weekday")
fillAct$dType <- as.factor(fillAct$dType)
```
**Make a time series plot of average number of steps taken, across weekday days or weekend days**    
  
```{r}
groupedAct <- aggregate(steps~interval + dType, fillAct, mean)
g <- qplot(interval, steps, data= groupedAct, geom= "line", facets= .~dType)
print(g)
```



